Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-01 Manuscript/manuscript.Rmd
editted abstract for BSRS

**Methods.** To investigate the problem of spatial genomic relationships and class imbalance problem, we developed an ensemble framework for optimizing prediction of TAD boundaries. We utilized several resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) coupled with various metrics for classifier performance evaluation (Area Under Receiver Operating Characteristics (AUCROC) curve, Precision-Recall curve (AUPRC), F1, Matthews Correlation Coefficient (MCC), among others) to select the best predictive model in class imbalance settings.

**Results.** Using different feature engineering and resampling strategies, coupled with elastic net-based feature selection, we systematically compared the performance of models predicting the location of TAD boundaries (Gm12878 and K562 Hi-C data at 10kb to 100kb resolutions). Precision-Recall curves were found to be best suited for evaluating model performance in class imbalance settings. Random undersampling and SMOTE-resampled data performed best for optimal parameter selection. Transcription Factor Binding Sites (TFBSs) were best type of genomic annotations for predicting TAD boundaries than histone modifications and chromatin states; we identified well-known CTCF, SMC3, RAD21, and several novel TFBSs as best predictors of TAD boundaries. Importantly, spatial relationships (distance between TAD boundaries and genomic annotations) were best predictors of TAD boundaries as compared with commonly used overlap counts or percent overlap measures.

>>>>

**Background.** Topologically Associated Domains (TADs) represent fundamental building blocks that lead to the organization and regulation of the three-dimensional (3D) genome. The enrichment of several epigenetic marks at TAD boundaries suggests a strong epigenetic component of TAD boundary formation. Several methods have been developed to leverage the growing amount of (epi)genomic annotation data for predicting TAD boundaries; however, they overlooked key characteristics of the data including spatial associations between genomic annotations and boundaries, as well as the severe imbalance in TAD-boundary vs non-TAD boundary defined genomic regions. Furthermore, resolution of Hi-C data remains well below the resolution of epigenomic assays (hundredths of bases for ChIP-seq) affecting the ability of Hi-C sequencing in discerning gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture.

**Methods.** We first developed an ensemble framework for optimizing data level characteristics necessary for TAD boundary prediction including combinations of 5 Hi-C data resolutions (5 kb to 100 kb), 3 spatially defined predictor types, and 4 re-sampling techniques, across the GM12878 and K562 cell lines. We then aimed to predict TAD boundary location on a 1D genomic map by validating optimally built cell line specific models at basepair resolution.

**Results.** We demonstrate with our ensemble framework that cell line specific random forest classifiers built on 10 kb resolution Hi-C data, with distance-type predictors, using SMOTE (*Synthetic Minority Oversampling Technique*) re-sampling yield optimal predictive performances, and outperform existing methods. Cell specific transcription factor binding sites (TFBS) were the best type of genomic annotations for predicting TAD boundaries when directly compared with histone modifications and chromatin states; we identified the well-known CTCF, SMC3, RAD21, and ZNF143 TFBS known to be enriched at TAD boundaries. Additionally, TAD boundary regions predicted at basepair resolution using our optimally built cell line specific models were found to be significantly closer to known enriched epigenomic marks indicating more accurate identification of TAD boundaries than existing Hi-C sequencing technologies.
